---
title: "Shiny apps prioritized"
author: "Wouter de Nooy"
date: "24 July 2017"
output: 
  html_document:
    number_sections : TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# regression-sampling

Section 8.1.5, Sampling distributions and assumptions.

```{r regression-sampling, fig.cap="What happens to regression lines from sample to sample?"}
# Goal: Understand that regression constant and coefficient(s) have sampling
# distributions.
# Generate a population with a weak negative effect (-0.6) of exposure on
# attitude and exposure, with a sizable error term (so a lot of variation in
# sample regression lines).
# Generate a sample (N = 10) and display it in a scatterplot with regression
# line, labelled with it's unstandardized regression coefficient value. Also
# plot the sampling distribution for the regression coefficient.
# Add a button to allow drawing a new sample; display the new sample and new
# regression line but retain the existing regression lines.
# Add button (or change sampling button) to draw 1,000 samples: don't display
# samples, just update sampling distribution with normal (or t) distribution as
# superimposed curve.
```

# interaction-var-effect

Section 8.3.2, Interaction variable.

```{r interaction-var-effect, fig.cap="What does an interaction variable do?"}
# Goal: Intuitive understanding of the effect of an interaction variable.
# Generate a dataset with 30 observations for the regression model y = 3 -
# 0.5x_1 + 1.5x_2 + 0.3x_1*x_2 with x_1 in the range [0, 10] and x_2 a dummy (0
# or 1) with a random uniform component to each parameter in the range [-.1,
# .1].
# In a scatterplot of attitude (Y) versus exposure (X), display the regression
# line (fat, grey) for the equation with x_2 = 0, labelled with the regression
# equation without the interaction variable. Display smokers and non-smokers
# with different colours/shapes.
# Add a select list labeled 'Add product of exposure and smoking status' with
# the values '--', '0 - Non-smokers', and '1 - Smokers'. Selection of a value
# adds the corresponding regression line to the plot with the category name and
# regression equation. (The line for non-smokers is parallel to the fat gray
# line.)
# Clicking/hovering over the newly created line shows the slope as a sum of the
# conditional and interaction effect, e.g., "Slope: 0.5 * exposure + 0.3 * 1 *
# exposure".
```

# symmetry-predictor-moderator

Section 8.5.4, Symmetry of predictor and moderator
Variant of mean-centering-moderator.

```{r symmetry-predictor-moderator, fig.cap=""}
# Goal: Understand the advantages of mean-centering the predictor by seeing how
# the reference value changes with mean-centering (and centering on another
# value, e.g., M plus/minus 1 SD).
# Use same data set as in app continuous-moderator: predictor = exposure,
# moderator = contact.
# Display scatterplot (x axis not labelled) with conditional regression effect
# for predictor (blue) at moderator value = 0 and conditional effect of
# moderator (red) for predictor = 0.
# Show two additional x axes marking the reference values of the predictor
# (blue) and moderator (red) (range [0, 10], initial value 0).
# Add sliders 'Exposure - x' and 'Contact - x' (equal length as two x axes,
# range [0, 10], initial value 0), labeled with values M - SD, M, and M + SD.
# Adjusting the sliders update the scale of the appropriate x axis (the marked
# point zero moves) and the regression lines in the scatterplot.
```

#	replacement

Section 2.1.1, Sampling with and without replacement.

```{r replacement, fig.cap="Sampling with and without replacement."}
#Create a sample consisting of 10 candies (dots) that are coloured (5 colours)
#and numbered. Display this sample as a dotplot.
# Add a button to create three (bootstrap) samples with replacement and a button
# to create three (bootstrap) samples without replacement.
# Show the three samples: coloured dots with their ID numbers.
```

#	bootstrap-ci

Section 3.5.3, Confidence intervals with bootstrapping
Variant of app bootstrapping.

```{r bootstrap-ci, eval=FALSE, echo=FALSE, fig.cap="How do we construct confidence intervals with bootstrapping?"}
# Adapt app bootstrapping: Show Initial Sample and Sampling Distribution,
# buttons Bootstrap 1000 samples and Draw new initial sample. Increase vertical
# size of sampling distribution but reduce range from 0 to 0.5.
# Add vertical lines for 95% confidence interval limits calculated from
# percentiles in the bootstrapped sampling distribution and limits calculated
# from the standard error (= standard deviation of the bootstrapped sampling
# distribution) times the critical value 1.96 (using a normal approximation).
# Finally, display the average of the bootstrapped sampling distribution as well
# as the true population proportion (0.2).
```

# reshyp-althyp

Section 5.5, 
Adapt type1vs2 or https://github.com/alice-i-cecile/power-calculator-toy? 

```{r reshyp-althyp, fig.cap="Sample size calculator."}
# Adapt app power-calculator-toy
# (https://github.com/alice-i-cecile/power-calculator-toy).
# Show sampling distribution (blue) for average candy weight according to
# research hypothesis (= null hypothesis): 2.8. Colour (blue) the area for not
# rejecting the null and add 95% as label.
# Add sampling distribution (red) for true average candy weight with slider to
# change true population average (range [2, 6], initial value 3.1. different
# from 2.8. Colour (red) areas above the rejection region and add labels with
# percentages for areas (rounded).
# Adjusting the slider changes the location of true sampling distribution, the
# size of the areas under it representing test power, and associated
# percentages.
```

