---
title: "Shiny apps prioritized"
author: "Wouter de Nooy"
date: "24 July 2017"
output: 
  html_document:
    number_sections : TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

#	regression-categorical

Section 8.1.4, A categorical predictor and dummy variables.

```{r regression-categorical, fig.cap="What are the predictive effects of smoking status?"}
# Goal: Understanding effects of dummy variables by manipulating the reference
# group.
# Generate numerical outcome data (attitude, range [-5, 5]) for a categorical
# predictor with 3 categories: (1) non-smoker, (2) former smoker, (3) smoker,
# and a random choice of one of the following scenarios:
#* (1 M = -1.7, SD = 1) < (2) = (3 M = 0.75, SD = 1) {initial situation},
#* (1 M = -1.7, SD = 1) = (2) < (3 M = 0.75, SD = 1), 
#* (2  M = -1.7, SD = 1) < (1) = (3 M = 0.75, SD = 1),
#* (1 M = -1.7, SD = 1) < (2 M = 0.75, SD = 1) < (3 M = 1.8, SD = 1). 
# Display jittered scatterplot containing the three groups with group means
# indicated by a line segment and value, regression lines through the reference
# group mean and each of the other group means. Display b and p value of
# regression weights with the regression lines.
# Add input to select the reference group, initially set to group (1). Update
# regression lines and their p values on selection of a new reference group.
# Add button to generate a new plot (with a new scenario).
```

# normal-approx-proportion

Section 2.5.2, Conditions for the use of theoretical probability distributions 
Simple variant of binomial histogram from bootstrapping with superimposed standard normal distribution from normal-approximation.

```{r normal-approx-proportion, fig.cap="How does the shape of the distribution of sample proportions change with sample size and proportion value?"}
#Generate a binomial sampling distribution with population proportion 0.5 and a
#large sample size and display it as a histogram.
# Project the normal distribution with population proportion p as mean and sqrt
# of p(1 - p)/N as standard deviation. Allow the user to change sample size and
# population proportion and update both the sampling distribution and the normal
# curve.
#perhaps adapt/simplify CLT_prop from Shiny-Ed on GitHub,
#https://github.com/ShinyEd/ShinyEd/tree/master/CLT_prop
```

# anova-twoway

Section 7.3, Different Score Levels for Two Factors
Adapt/extend anova-between.

```{r anova-twoway, fig.cap="?"}
# Goal: Illustrate that different main effects merely use means of different
# groupings.
# Similar to app anova-between but with a double classification of cases
# (according to endorser and sex) and the option to display means of different
# groupings.
# Generate 6 sets of 2 (random) observations (from a normally distributed
# population with mean runif(3, 7) and sd = 1). Assign the groups to the
# experimental treatment factor (endorser, 3 levels) and sex factor (2 levels).
# Represent observations in a dotplot with treatment as dot colour and sex as
# dot shape, each observation with a separate value on the x axis, ordered by
# factor levels.
# Display the grand mean as a horizontal line. 
# Allow the user to select (display) group means on one of the two factors or
# both. On selection, add group means as horizontal lines (coloured for the 3
# levels factor and different line styles for 2 levels factor) with between
# group variation indicated by double-sided arcs between group mean and grand
# mean for each dot (between). Add subgroup (endorser by sex) means if both
# factors are selected.
```

# regression-dichotomy

Section 8.1.3, Dichotomous predictors.

```{r regression-dichotomy, eval=FALSE, fig.cap="What is the difference in attitude between non-smokers and smokers?"}
# Goal: Refresh interpretation of unstandardized regression weight for a
# dichotomous predictor by manipulating group averages.
# Generate attitude scores  with average values -0.6 for non-smokers and 1.0 for
# smokers (N = 20 per group).
# Draw horizontal and vertical lines from the axes to the group means and add a
# regression line (line through the two group means).
# Display the current regression equation beneath or in the plot. 
# Allow the user to change the average score per group. Update the scatterplot,
# regression line and equation, and the horizontal/vertical lines.
# Generate predictor.
x <- c(rep(0, 20), rep(1, 20))
# Generate random errors per group.
set.seed(1272)
e <- rnorm(20, mean = 0, sd = 1)
e <- sign(e) * sqrt(abs(e*abs(e) - mean(e*abs(e)))) # center in squares
e <- c(e, e)
# Set group means
nonsmoker <- -0.6
smoker <- 1.0
# Generate outcome scaled by unstandardized regression coefficient.
y <- nonsmoker + smoker*x + e
lm(y~x)
```

# regression-sampling

Section 8.1.5, Sampling distributions and assumptions.

```{r regression-sampling, fig.cap="What happens to regression lines from sample to sample?"}
# Goal: Understand that regression constant and coefficient(s) have sampling
# distributions.
# Generate a population with a weak negative effect (-0.6) of exposure on
# attitude and exposure, with a sizable error term (so a lot of variation in
# sample regression lines).
# Generate a sample (N = 10) and display it in a scatterplot with regression
# line, labelled with it's unstandardized regression coefficient value. Also
# plot the sampling distribution for the regression coefficient.
# Add a button to allow drawing a new sample; display the new sample and new
# regression line but retain the existing regression lines.
# Add button (or change sampling button) to draw 1,000 samples: don't display
# samples, just update sampling distribution with normal (or t) distribution as
# superimposed curve.
```

# dichotomous-moderator

Section 8.3.1, A dichotomous moderator and continuous predictor.

```{r dichotomous-moderator, fig.cap="Is the effect of exposure on attitude moderated by smoking status?"}
# Goal: Sensitize the student to the notion that moderation in a regression
# model means different slopes for different groups.
# In a graph with attitude as Y axis and exposure as X axis, generate two regression lines, one for smokers and one for non-smokers. 
# Plot the regression lines in a scatterplot. 
# Show the regression equation for each line (preferably in the plot). 
# Systematically vary the slopes (the same, one more negative than the other,
# opposite signs) and the intercept difference (positive, nearly zero,
# negative).
# Add a Generate New button to replace the regression lines by a new pair of lines.
```

# interaction-var-effect

Section 8.3.2, Interaction variable.

```{r interaction-var-effect, fig.cap="What does an interaction variable do?"}
# Goal: Intuitive understanding of the effect of an interaction variable.
# Generate a dataset with 30 observations for the regression model y = 3 -
# 0.5x_1 + 1.5x_2 + 0.3x_1*x_2 with x_1 in the range [0, 10] and x_2 a dummy (0
# or 1) with a random uniform component to each parameter in the range [-.1,
# .1].
# In a scatterplot of attitude (Y) versus exposure (X), display the regression
# line (fat, grey) for the equation with x_2 = 0, labelled with the regression
# equation without the interaction variable. Display smokers and non-smokers
# with different colours/shapes.
# Add a select list labeled 'Add product of exposure and smoking status' with
# the values '--', '0 - Non-smokers', and '1 - Smokers'. Selection of a value
# adds the corresponding regression line to the plot with the category name and
# regression equation. (The line for non-smokers is parallel to the fat gray
# line.)
# Clicking/hovering over the newly created line shows the slope as a sum of the
# conditional and interaction effect, e.g., "Slope: 0.5 * exposure + 0.3 * 1 *
# exposure".
```

# symmetry-predictor-moderator

Section 8.5.4, Symmetry of predictor and moderator
Variant of mean-centering-moderator.

```{r symmetry-predictor-moderator, fig.cap=""}
# Goal: Understand the advantages of mean-centering the predictor by seeing how
# the reference value changes with mean-centering (and centering on another
# value, e.g., M plus/minus 1 SD).
# Use same data set as in app continuous-moderator: predictor = exposure,
# moderator = contact.
# Display scatterplot (x axis not labelled) with conditional regression effect
# for predictor (blue) at moderator value = 0 and conditional effect of
# moderator (red) for predictor = 0.
# Show two additional x axes marking the reference values of the predictor
# (blue) and moderator (red) (range [0, 10], initial value 0).
# Add sliders 'Exposure - x' and 'Contact - x' (equal length as two x axes,
# range [0, 10], initial value 0), labeled with values M - SD, M, and M + SD.
# Adjusting the sliders update the scale of the appropriate x axis (the marked
# point zero moves) and the regression lines in the scatterplot.
```

#	replacement

Section 2.1.1, Sampling with and without replacement.

```{r replacement, fig.cap="Sampling with and without replacement."}
#Create a sample consisting of 10 candies (dots) that are coloured (5 colours)
#and numbered. Display this sample as a dotplot.
# Add a button to create three (bootstrap) samples with replacement and a button
# to create three (bootstrap) samples without replacement.
# Show the three samples: coloured dots with their ID numbers.
```

#	bootstrap-ci

Section 3.5.3, Confidence intervals with bootstrapping
Variant of app bootstrapping.

```{r bootstrap-ci, eval=FALSE, echo=FALSE, fig.cap="How do we construct confidence intervals with bootstrapping?"}
# Adapt app bootstrapping: Show Initial Sample and Sampling Distribution,
# buttons Bootstrap 1000 samples and Draw new initial sample. Increase vertical
# size of sampling distribution but reduce range from 0 to 0.5.
# Add vertical lines for 95% confidence interval limits calculated from
# percentiles in the bootstrapped sampling distribution and limits calculated
# from the standard error (= standard deviation of the bootstrapped sampling
# distribution) times the critical value 1.96 (using a normal approximation).
# Finally, display the average of the bootstrapped sampling distribution as well
# as the true population proportion (0.2).
```

# reshyp-althyp

Section 5.5, 
Adapt type1vs2 or https://github.com/alice-i-cecile/power-calculator-toy? 

```{r reshyp-althyp, fig.cap="Sample size calculator."}
# Adapt app power-calculator-toy
# (https://github.com/alice-i-cecile/power-calculator-toy).
# Show sampling distribution (blue) for average candy weight according to
# research hypothesis (= null hypothesis): 2.8. Colour (blue) the area for not
# rejecting the null and add 95% as label.
# Add sampling distribution (red) for true average candy weight with slider to
# change true population average (range [2, 6], initial value 3.1. different
# from 2.8. Colour (red) areas above the rejection region and add labels with
# percentages for areas (rounded).
# Adjusting the slider changes the location of true sampling distribution, the
# size of the areas under it representing test power, and associated
# percentages.
```

